### 2021-07-07 ch3bedEle.tex
rolling on the conclusion
this is a nightmare

\chapter{Analysis of bed elevation change and sediment transport fluctuations}
\label{ch:bedEle}

The transport characteristics of coarse grains moving under a turbulent flow ultimately control a wide set processes within rivers, including the export of contaminants \citep{Malmon2005,Macklin2006}, the success of ecological restoration efforts \citep{Gaeuman2017}, and the response of channel morphology to disturbances \citep{Hassan2017}.
Although the displacements of individual grains are certainly a mechanical consequence of forces imparted from the flow, bed, and other grains \citep{Wiberg1985, Vowinckel2014,Gonzalez2017}, accurately characterizing these forces within natural channels is practically impossible, especially considering the intense variability these forces display \citep{Schmeeckle2007,Celik2010, Dwivedi2011}. In response, investigators have developed a stochastic concept of bedload transport \citep{Einstein1937}, whereby the erosion and deposition of individual grains are modeled as the random results of undetermined forces \citep{Einstein1950,Paintal1971,Ancey2006}.

Essentially two types of bedload transport model have been developed from this concept. 
The first type provides the probabilistic dynamics of a small population of tracer grains as they transport downstream \citep{Einstein1937,Hubbell1964, Nakagawa1976,Martin2012,Lajeunesse2017,Wu2019}, while the second provides the statistics of the number of moving grains (``the particle activity'') within a control volume \citep{Einstein1950,Ancey2006,Ancey2008,Furbish2012a}.
In the first type, individual displacements are considered to result from alternate step-rest sequences, where step lengths and resting times are random variables following statistical distributions \citep{Einstein1937}. 
Differences between the random-walk motions of one grain and the next imply a spreading apart of tracer grains as they transport downstream: bedload tracers undergo diffusion.

Resting time distributions have been carefully studied in relation to these models because the predicted diffusion characteristics are critically dependent on whether the distribution has a light or heavy tail \citep{Bradley2017,Martin2012,Weeks1998}.
Resting times have puzzled researchers because early experiments show exponential distributions \citep{Einstein1937,Hubbell1964, Yano1969,Nakagawa1976}, while later experiments show heavy-tailed power-law distributions \citep{Martin2012,Voepel2013, Olinde2015, Pretzlav2016a, Bradley2017, Liu2019}.
A predominant hypothesis is that power-law distributed resting times originate from buried grains \citep{Martin2014,Voepel2013}; this hypothesis permits surface grains to retain exponential resting times. 
Conceptually, when grains rest on the surface, material transported from upstream can deposit on top of them, preventing entrainment until its removal, driving up resting times and imparting a heavy tail to the distribution.
To our knowledge, \citet{Martin2014} have provided the only direct support for this hypothesis by tracking grains through complete cycles of burial and exhumation using a narrow flume with glass walls.
They observed heavy-tailed resting times of buried grains and described their results with a mathematical model similar to an earlier effort by \citet{Voepel2013}.
Both of these models treat bed elevation changes as a random walk and interpret resting times as return periods from above in the bed elevation time-series \citep{Redner2007}. Each describes resting time distributions from different experiments, but they rely on different random walk models, and their treatment of bed elevations as a process independent of sediment transport is questionable at first glance, since bedload transport is the source of bed elevation changes \citep{Wong2007}, and neither model explicitly includes bedload transport.
Models of sedimentary bed evolution incorporating sediment transport processes might enhance understanding of sediment resting times.

The second type of stochastic model prescribes rates (probabilities per unit time) to the erosion and deposition events of individual grains within a control volume to calculate the particle activity \citep{Einstein1950}.
These approaches aim at a complete statistical characterization of the bedload flux \citep{Furbish2012a,Fathel2015,Furbish2017,Heyman2016}, including probability distributions \citep{Ancey2006,Ancey2008}, spatial and temporal characteristics of its fluctuations \citep{Ancey2008, Heyman2014a, Roseberry2012,Dhont2018}, and the dependence of these statistical characteristics on the length and time scales over which they are measured or calculated \citep{Singh2009,Singh2012,Saletti2015}.
A recent surge in research activity has generated rapid progress and spawned many new inquiries in this subject. 
For example, \citet{Ancey2006} demonstrated that a constant erosion rate as originally proposed by \citet{Einstein1950} was insufficient to develop realistically large particle activity fluctuations, so they added a positive feedback between the particle activity and erosion rate they called ``collective entrainment'' \citep{Ancey2008, Heyman2013,Heyman2014,Ma2014, Lee2018}.
While they deemed this feedback necessary to model realistic activity fluctuations, the implications of this collective entrainment term on bed topography and particle activity changes has not been fully explored.

In this work, we present the first stochastic model coupling the erosion and deposition of individual bedload grains to local bed elevation changes.
Our model extends the \citet{Ancey2008} model to describe the interplay between bedload flux and bed elevation fluctuations in a control volume.
This development permits a systematic study of the repercussions of collective entrainment, and it frames bed elevation changes as a direct consequence of the sediment transport process.
Our model has two key assumptions: (1) bedload erosion and deposition can be characterized by probabilities per unit time, or rates \citep{Einstein1950, Ancey2008}; and (2) these rates are contingent on the local bed elevation, encoding the property that erosion of sediment is emphasized from regions of exposure, while deposition is emphasized in regions of shelter \citep{Sawai1987, Wong2007}.
We study statistical characteristics of bedload transport, bed elevation, and resting times of sediment undergoing burial using a mixture of numerical simulations and analytical approximations. We introduce the stochastic model in section \ref{sec:model}, and we solve it in section \ref{sec:solution} with a mixture of numerical and analytical techniques. We discern several new features of particle activity and bed elevation statistics that result from feedbacks between the erosion and deposition rates and the local bed elevation. We present these features in section \ref{sec:results}. We conclude with the implications of our results and speculate on topics for future research in sections \ref{sec:discussion} and \ref{sec:conclusion}.

\section{Stochastic model of bedload transport and bed elevations}
\label{sec:model}

We prescribe a volume of downstream length $L$ containing some number $n$ of moving particles in the flow and some number $m$ of stationary particles composing the bed at time $t$, as depicted in figure \ref{fig:definition}. We define $m$ relative to the mean number of grains within the control volume, so that it can be either positive or negative. $n$ is always a positive integer including $0$.
For simplicity, we consider all particles as approximately spherical with the same diameter $2a$, so their mobility and packing characteristics are consistent.
Following \citet{Ancey2008}, we prescribe four events that can occur at any instant to modify the populations $n$ and $m$, and we characterize these events using probabilities per unit time (rates).
These events are (1) migration of a moving particle into the volume from upstream ($n \rightarrow n+1$), (2) the entrainment (erosion) of a stationary particle into motion within the volume ($m\rightarrow m-1$ and $n\rightarrow n+1$), (3) the deposition of a moving particle to rest within the volume ($m\rightarrow m+1$ and $n\rightarrow n-1$), and (4) the migration of a moving particle out of the volume to downstream ($n\rightarrow n-1$).
The four events are depicted as arrows in figure \ref{fig:definition}.
As the events occur at random intervals, they set up a joint stochastic evolution of the populations $n$ and $m$ characterized by a joint probability distribution $P(n,m,t)$ for the number of particles in motion and rest in the volume at $t$.
\begin{figure}[!htbp]
	\includegraphics[width=\linewidth,keepaspectratio]{./figures/ch3/definition-combo.pdf}
	\caption{Definition sketch of a control volume containing $n$ moving grains and $m$ resting grains. Migration, entrainment, and deposition are represented by arrows, and the instantaneous bed elevation is depicted by dotted lines. The bed is displayed in a degraded state, where $m<0$. The marginal distributions of $n$ and $m$ are indicated in the upper right panel, while the bottom panel is a realized time-series of bed elevations computed from $m$ using (\ref{eq:ele}).}
	\label{fig:definition}
\end{figure}
The populations $n$ and $m$ provide the bulk bedload flux $q_s$ and the local bed elevation $z$.
The mean bedload transport rate is given by $q_s = u_s\langle n \rangle/L$, where $u_s$ is the characteristic velocity of moving bedload and $\langle n \rangle = 
\sum_{n,m}nP(n,m) $ is the mean number of grains in motion \citep{Charru2004, Ancey2008, Furbish2012a}.
The bed elevation is related to $m$ through the packing geometry of the bed.
To quantify this, we introduce a packing fraction $\phi$ of grains in the bed \citep{Bennett1972}, and for simplicity we consider the bed as two-dimensional \citep{Einstein1950, Paintal1971}. The deviation from the mean bed elevation is then
\begin{equation} z(m) = \frac{\pi a^2}{\phi L}m = z_1 m. \label{eq:ele}\end{equation}
The constant $z_1 = \pi a^2/(\phi L)$ is an important scale of the problem. 
$z_1$ is the magnitude of bed elevation change in an average sense across the control volume associated with the addition or removal of a single grain.

Bed elevation changes modify the likelihood of entrainment and deposition in a negative feedback \citep{Sawai1987, Wong2007}; that is, aggradation increases the likelihood of entrainment, while degradation increases the likelihood of deposition.
\citet{Wong2007} concluded that bed elevation changes induce an exponential variation in entrainment and deposition probabilities, while \citet{Sawai1987} concluded that the variation is linear.
For simplicity, we incorporate the scaling of \citet{Sawai1987} and note its equivalence to the \citet{Wong2007} scaling when bed elevation changes are small.
Because experimental distributions of bed elevations are often symmetrical, \citep{Crickmore1962, Pender2001,Wong2007, Martin2014}, we expect the erosion and deposition feedbacks to have the same strength.
That is, as bed elevation changes drive up (down) erosion rates, so they drive down (up) deposition rates to the same degree.
Merging these ideas with those of \citet{Ancey2008}, we write the four possible transitions with local bed elevation-dependent entrainment and deposition rates as
\begin{align}
	&R_{MI}(n+1|n) = \nu && &\text{migration in}, \label{eq:rate1}\\
	&R_E(n+1,m-1|n,m)=(\lambda + \mu n)[1 + \kappa m], && &\text{entrainment}, \label{eq:rate2}\\
	&R_D(n-1,m+1|n,m)=\sigma n [1- \kappa m ], && &\text{deposition}. \label{eq:rate3} \\
	&R_{MO}(n-1|n) =\gamma n && &\text{migration out} \label{eq:rate4}.
\end{align}
In equations (\ref{eq:rate2}) and (\ref{eq:rate3}), $\kappa$ is a coupling constant between bed elevations and the entrainment and deposition rates.
$\nu$ is the rate of migration into the control volume, $\lambda$ is the conventional entrainment rate, $\mu$ is the collective entrainment rate, $\sigma$ is the deposition rate, and $\gamma$ is the rate of migration out of the control volume.
At $m=0$, these equations reduce to those of the \citet{Ancey2008} model.
Away from this elevation, entrainment and deposition are alternatively suppressed and enhanced depending on the sign of $m$, constituting a feedback between bed elevation changes and erosion and deposition.
We refer to $\kappa$ as a coupling constant since it controls the strength of this feedback. We later demonstrate the relationship
\begin{equation}\kappa \approx \big(\frac{z_1}{2l}\big)^2 \label{eq:active}
\end{equation}
where $l$ is a characteristic length scale of bed elevation change that we interpret as the active layer depth \citep{Wong2007,Church2017}.
All four rates are independent of the past history of the populations and depend only on the current populations $(n,m)$. 
As a result, the model is Markovian \citep{Cox1965, VanKampen2007}, meaning time intervals between any two subsequent transitions are exponentially distributed \citep{Gillespie2007}.

We write the master equation for the probability flow using the forward Kolmogorov equation $\partial P(n,m;t)/\partial t = 
\sum_{n',m'} [R(n,m|n',m')P(n',m';t)-R(n',m'|n,m)P(n,m;t)]$ \citep{Cox1965, Gillespie1991, Ancey2008} as 
\begin{multline}
	\frac{\partial P}{\partial t}(n,m;t) =  
	\nu P(n-1,m;t) + [\lambda(m+1) + \mu(n-1)][1+\kappa(m+1)]P(n-1,m+1;t)\\  
	+ \sigma(n+1)[1-\kappa(m-1)]P(n+1,m-1;t) + \gamma(n+1) P(n+1,m;t) \\
	- 
	\{ \nu + \lambda+ \mu n (1+\kappa m) +  \sigma n ( 1- \kappa m) + \gamma n \}P(n,m;t).
	\label{eq:master}
\end{multline}
The joint probability distribution $P(n,m;t)$ solving this equation fully characterizes the statistics of $n$ and $m$ -- proxies for the bedload flux and local bed elevation.
The average entrainment and deposition rates $E$ and $D$ over all bed elevations are $E = \lambda +\mu \langle n \rangle$ and $D=\sigma \langle n \rangle$.
We anticipate that solutions of (\ref{eq:master}) will adjust from the initial conditions to a steady-state distribution $P_s(n,m)$ -- independent of time -- if the constant factors in the transition rates are representative of equilibrium conditions. Equilibrium requires $E=D$, meaning there is no net change in elevation, and $\nu = \gamma \langle n \rangle$, meaning mass is conserved in the control volume (inflow $=$ outflow).
This Master equation describes a two-species stochastic birth-death model \citep{Cox1965} of a type well-known in population ecology \citep{Pielou1977, Swift2002} and chemical physics \citep{Gardiner1983}.
In our context, the two populations are the moving and stationary grains in the volume.

\section{Model solutions}
\label{sec:solution}

Unfortunately, equation (\ref{eq:master}) does not appear to admit an analytical solution unless $\kappa=0$ (but see \citet{Swift2002} for the generating function method which fails in this case).
The difficulty originates from the product terms between $n$ and $m$ representing the bed elevation dependence of collective entrainment and deposition rates.
In response to this difficulty, we resort to numerical methods and analytical approximations, simulating equation (\ref{eq:master}) with the Gillespie algorithm \citep{Gillespie1977, Gillespie1991, Gillespie2007} and solving it approximately with mean field and Fokker-Planck approaches \citep{Haken1978,Gardiner1983}. The simulation algorithm is described in \ref{sec:numerical}, and analytical approximations are described in \ref{sec:analytical}.

@@

The Gillespie algorithm leverages the defining property of a Markov process: when transition rates are independent of history, time intervals between transitions are \begin{wraptable}{l}{0.5\textwidth}
	\caption{Migration, entrainment, and deposition rates at $z(m)=0$ from \citet{Ancey2008}. Units are $s^{-1}$ (probability/time). In our model, bed elevation changes modulate these rates in accord with (\ref{eq:rate1}-\ref{eq:rate4}).}\label{tab:anceyparams}
	\begin{tabular}{cccccc} \\ 
		\toprule  
		flow & $\nu$ & $\lambda$ & $\mu$ & $\sigma$ & $\gamma$ \\
		\midrule
		(a) & 5.45  & 6.59  & 3.74 & 4.67 & 0.77 \\
		\midrule
		(g) & 7.74  & 8.42  & 4.34 & 4.95 & 0.56 \\
		\midrule
		(i) & 15.56 & 22.07 & 3.56 & 4.52 & 0.68 \\
		\midrule
		(l) & 15.52 & 14.64 & 4.32 & 4.77 & 0.48 \\
		\bottomrule
	\end{tabular}
\end{wraptable}exponentially distributed \citep{Cox1965}.
As a result, to step the Markov process through a single transition, one can draw a first random value from the exponential distribution of transition intervals to determine the time of the next transition,
then draw a second random value to choose the type of transition that occurs using relative probabilities formed from equations (\ref{eq:rate1}-\ref{eq:rate4}). The transition is enacted by shifting $t$, $n$ and $m$ by the appropriate values to the type of transition (that is, entrainment is $m\rightarrow m-1$ and $n \rightarrow n+1$, and so on).
This procedure can be iterated to form an exact realization of the stochastic process \citep{Gillespie2007}.
We provide additional background on the stochastic simulation method in the supplementary material and refer the reader to \citet{Gillespie2007} for more detail.

Using this method, we simulated 4 transport conditions with 13 different values of $l$ taken across a range from $l=a$ (a single radius) to $l=10a$ (10 radii).
These values include the range exhibited by the available experimental data on bed elevation timeseries \citep{Wong2007,Singh2009,Martin2014}.
For the migration, entrainment, and deposition parameters representing bedload transport at each flow condition $(\nu, \lambda, \mu, \sigma, \gamma)$, we used the values measured by \citet{Ancey2008} in a series of flume experiments: these are summarized in table \ref{tab:anceyparams}.
Flow conditions are labeled (a), (g), (i), and (l), roughly in order of increasing bedload flux (see \citet{Ancey2008} for more details). 
In all simulations, we take the packing fraction $\phi = 0.6$ -- a typical value for a pile of spheres \citep[e.g.,][]{Bennett1972}, and we set $L = 22.5$cm and $a = 0.3$cm in accord with the \citet{Ancey2008} experiments.
Each simulation was run for $250$ hours of virtual time, a period selected to ensure neat convergence of particle activity and bed elevation statistics.

@@

\section{Discussion}
\label{sec:discussion}

@@

In this paper, we have built on earlier works \citep{Ancey2008,Martin2014} to include the vertical dimension of bed elevation dynamics, study the interplay between bedload transport and bed elevation fluctuations, and investigate resting time distributions of sediment undergoing burial.
To our knowledge, this model is the first description of bedload transport and bed elevations as a coupled stochastic population model based on individual grains.
Numerical solutions and analytical approximations provided negative binomial distributions of bedload activity and normal distributions of bed elevations.
Although experiments under more natural conditions with segregation processes,  migrating bedforms, or sediment supply perturbations have shown particle activity distributions with heavier tails \citep{Dhont2018,Saletti2015} and non-Gaussian bed elevations \citep{Singh2012,Aberle2006}, our results reproduce the key features of the most controlled bed elevation \citep{Wong2007,Martin2014} and bedload transport \citep{Heyman2016,Ancey2008} experiments in the literature.

>>>>

\chapter{Analysis of bed elevation change and sediment transport fluctuations}
\label{ch:ch3}

The transport characteristics of coarse grains moving under a turbulent flow ultimately control a wide set processes within rivers, including the export of contaminants \citep{Malmon2005,Macklin2006}, the success of ecological restoration efforts \citep{Gaeuman2017}, and the response of channel morphology to disturbances \citep{Hassan2017}.
Although the displacements of individual grains are a mechanical consequence of forces imparted from the flow, bed, and other grains \citep{Wiberg1985, Vowinckel2014,Gonzalez2017}, accurately characterizing these forces within natural channels is practically impossible, especially considering the intense variability these forces display \citep{Schmeeckle2007,Celik2010, Dwivedi2011}.
In response, researchers have developed a stochastic concepts of bedload transport \citep{Einstein1937}, whereby the erosion and deposition of individual grains are modelled as the random results of undetermined forces \citep{Einstein1950,Paintal1971,Ancey2006}.

Essentially two types of bedload transport model have been developed from this concept.
The first type provides the probabilistic dynamics of a small population of tracer grains as they transport downstream \citep{Einstein1937,Hubbell1964, Nakagawa1976,Martin2012,Lajeunesse2017,Wu2019}, while the second provides the statistics of the number of moving grains (``the particle activity'') within a control volume \citep{Einstein1950,Ancey2006,Furbish2012a}.
In the first type, individual displacements are considered to result from alternate step-rest sequences, where step lengths and resting times are random variables following statistical distributions \citep{Einstein1937}. 
Differences between the random-walk motions of one grain and the next imply a spreading apart of tracer grains as they transport downstream: bedload tracers undergo diffusion.

Resting time distributions have been carefully studied in relation to these models because the predicted diffusion characteristics are critically dependent on whether the distribution has a light or heavy tail \citep{Bradley2017,Martin2012,Weeks1998}.
Resting times have puzzled researchers because early experiments showed exponential distributions \citep{Einstein1937,Hubbell1964, Yano1969,Nakagawa1976}, while later experiments showed heavy-tailed power-law distributions \citep{Martin2012,Voepel2013, Olinde2015, Pretzlav2016a, Bradley2017, Liu2019}.
A predominant hypothesis is that power-law distributed resting times originate from buried grains \citep{Martin2014,Voepel2013}. This permits surface grains to retain exponential resting times. 
Conceptually, when grains rest on the surface, material transported from upstream can deposit on top of them, preventing entrainment until its removal, driving up resting times and imparting a heavy tail to the distribution.

\citet{Martin2014} have provided the only direct support for this hypothesis by tracking grains through complete cycles of burial and exhumation in a narrow flume.
They observed heavy-tailed resting times of buried grains and described their results with a mathematical model similar to an earlier effort by \citet{Voepel2013}.
Both of these models treat bed elevation changes as a random walk and interpret resting times as return periods from above in the bed elevation time-series \citep{Redner2007}.
Each describes resting time distributions from different experiments, but they rely on different random walk models, and their treatment of bed elevations as a process independent of sediment transport is questionable at first glance, since bedload transport is the source of bed elevation changes \citep{Wong2007}, while neither model explicitly describes bedload transport.
Models of sedimentary bed evolution incorporating sediment transport processes might enhance understanding of sediment resting times.

The second type of stochastic model prescribes rates (probabilities per unit time) to the erosion and deposition events of individual grains within a control volume to calculate the particle activity \citep{Einstein1950}.
These approaches aim at a complete statistical characterization of the bedload flux \citep{Furbish2012a,Fathel2015,Furbish2017,Heyman2016}, including probability distributions \citep{Ancey2006,Ancey2008}, spatial and temporal characteristics of its fluctuations \citep{Heyman2014a, Roseberry2012,Dhont2018}, and the dependence of these statistical characteristics on the length and time scales over which they are observed \citep{Singh2009,Singh2012,Saletti2015}.
A recent surge in research activity has generated rapid progress in this subject. 
For example, \citet{Ancey2006} demonstrated that a constant erosion rate as originally proposed by \citet{Einstein1950} was insufficient to develop realistically large particle activity fluctuations, so they added a positive feedback between the particle activity and erosion rate they called ``collective entrainment'' \citep{Ancey2008, Heyman2013,Heyman2014, Lee2018}.
While this feedback was deemed necessary to model realistic activity fluctuations, the implications of collective entrainment for the bed topography have not been fully explored.

In this work, I present the first stochastic model coupling the erosion and deposition of individual bedload grains to local bed elevation changes.
This model extends the birth-death model approach reviewed in section \ref{sec:elebirthdeath} to describe the interplay between the bedload flux and the bed elevation within a control volume.
This new model evaluates the effect of collective entrainment on bed elevation change, and it allows for investigations of the residence times of particles buried in the bed subsurface.
The model has two major assumptions: First, bedload erosion and deposition are characterized by probabilities per unit time, or rates \citep{Einstein1950, Ancey2008}; and second, these rates are contingent on the local bed elevation, encoding the property that erosion of sediment is emphasized from regions of exposure, while deposition is emphasized in regions of shelter \citep{Sawai1987, Wong2007}.

Below, section \ref{sec:elemodel} introduces the stochastic model and section \ref{sec:elesolution} solves it with a mixture of numerical and analytical techniques. Model solutions illustrate several new features of particle activity and bed elevation statistics that result from feedbacks between the erosion and deposition rates with bed elevation change. Section \ref{sec:eleresults} present these features, while sections \ref{sec:elediscussion} and \ref{sec:eleconclusion} conclude with the implications of these features and the suggestions they raise for future research.

\section{Stochastic model of bedload transport and bed elevations}
\label{sec:elemodel}

The model in this chapter is formulated by considering a volume of downstream length $L$ containing some number $n$ of moving particles in the flow and some number $m$ of stationary particles composing the bed at time $t$, as depicted in figure \ref{fig:eledefinition}. The value $m$ is defined relative to the mean number of grains within the control volume, so that it can be either positive or negative. $n$ is always a positive integer including $0$.

All particles are considered approximately spherical with the same diameter $2a$, so their mobility and packing characteristics are consistent.
Following \citet{Ancey2008}, basically four events can occur at any instant to modify the populations $n$ and $m$. These events are characterized using probabilities per unit time (rates).
These events are (1) migration of a moving particle into the volume from upstream ($n \rightarrow n+1$), (2) the entrainment (erosion) of a stationary particle into motion within the volume ($m\rightarrow m-1$ and $n\rightarrow n+1$), (3) the deposition of a moving particle to rest within the volume ($m\rightarrow m+1$ and $n\rightarrow n-1$), and (4) the migration of a moving particle out of the volume to downstream ($n\rightarrow n-1$).
The four events are depicted as arrows in figure \ref{fig:eledefinition}.
As the events occur at random intervals, they set up a joint stochastic evolution of the populations $n$ and $m$ characterized by a joint probability distribution $P(n,m,t)$ for the number of particles in motion and rest in the volume at $t$.
\begin{figure}[!htbp]
	\includegraphics[width=\linewidth,keepaspectratio]{./figures/ch3/definition-combo.pdf}
	\caption{Definition sketch of a control volume containing $n$ moving grains and $m$ resting grains. Migration, entrainment, and deposition are represented by arrows, and the instantaneous bed elevation is depicted by dotted lines. The bed is displayed in a degraded state, where $m<0$. The marginal distributions of $n$ and $m$ are indicated in the upper right panel, while the bottom panel is a realized time-series of bed elevations computed from $m$ using (\ref{eq:ele}).}
	\label{fig:eledefinition}
\end{figure}

The populations $n$ and $m$ provide the volumetric bedload flux $\Phi$ and the local bed elevation $z$.
The mean transport rate is given by $\Phi = u_s\langle n \rangle/L$, where $u_s$ is the characteristic velocity of moving bedload and $\langle n \rangle = 
\sum_{n,m}nP(n,m) $ is the mean number of grains in motion \citep{Charru2004, Ancey2008, Furbish2012a}.
The bed elevation is related to $m$ through the packing geometry of the bed.
This relationship depends on the packing fraction $\phi$ of grains in the bed \citep{Bennett1972}. Considering the bed as two-dimensional \citep{Einstein1950, Paintal1971}, the deviation from the mean bed elevation can be expressed as
\begin{equation} z(m) = \frac{\pi a^2}{\phi L}m = z_1 m. \label{eq:ele}\end{equation}
The constant $z_1 = \pi a^2/(\phi L)$ is an important scale in the problem. 
$z_1$ is the magnitude of bed elevation change in an average sense across the control volume associated with the addition or removal of a single grain.

Bed elevation changes modify the likelihood of entrainment and deposition in a negative feedback \citep{Sawai1987, Wong2007}.
Aggradation increases the likelihood of entrainment, while degradation increases the likelihood of deposition.
\citet{Wong2007} concluded that bed elevation changes induce an exponential variation in entrainment and deposition probabilities, while \citet{Sawai1987} concluded that the variation is linear.
For simplicity, this chapter utilizes the scaling of \citet{Sawai1987}. This scaling is equivalent to the  \citet{Wong2007} scaling when bed elevation changes are small.
Because experimental distributions of bed elevations are often symmetrical, \citep{Crickmore1962, Pender2001,Wong2007, Martin2014}, the erosion and deposition feedbacks are considered to have the same strength.
As bed elevation changes drive up (down) erosion rates, so they drive down (up) deposition rates to the same degree.

Merging these ideas with those of \citet{Ancey2008} produces expressions for the four possible transitions with local bed elevation-dependent entrainment and deposition rates:
\begin{align}
	&R_{MI}(n+1|n) = \nu && &\text{migration in}, \label{eq:rate1}\\
	&R_E(n+1,m-1|n,m)=(\lambda + \mu n)[1 + \kappa m], && &\text{entrainment}, \label{eq:rate2}\\
	&R_D(n-1,m+1|n,m)=\sigma n [1- \kappa m ], && &\text{deposition}. \label{eq:rate3} \\
	&R_{MO}(n-1|n) =\gamma n && &\text{migration out} \label{eq:rate4}.
\end{align}
In equations (\ref{eq:rate2}) and (\ref{eq:rate3}), $\kappa$ is a coupling constant between bed elevations and the entrainment and deposition rates.
$\nu$ is the rate of migration into the control volume, $\lambda$ is the conventional entrainment rate, $\mu$ is the collective entrainment rate, $\sigma$ is the deposition rate, and $\gamma$ is the rate of migration out of the control volume.
At $m=0$, these equations reduce to those of the \citet{Ancey2008} model.
Away from this elevation, entrainment and deposition are alternatively suppressed or enhanced depending on the sign of $m$, constituting a feedback between bed elevations and erosion / deposition.

All four rates are independent of the past history of the populations and depend only on the current populations $(n,m)$. 
As a result, the model is Markovian \citep{Cox1965, VanKampen2007}, meaning time intervals between any two subsequent transitions are exponentially distributed \citep{Gillespie2007}.

The master equation for the probability flow can be written using the Komogorov equation
 $\partial P(n,m;t)/\partial t = 
\sum_{n',m'} [R(n,m|n',m')P(n',m';t)-R(n',m'|n,m)P(n,m;t)]$ \citep{Cox1965, Gillespie1991, Ancey2008} as 
\begin{multline}
	\frac{\partial P}{\partial t}(n,m;t) =  
	\nu P(n-1,m;t) + [\lambda(m+1) + \mu(n-1)][1+\kappa(m+1)]P(n-1,m+1;t)\\  
	+ \sigma(n+1)[1-\kappa(m-1)]P(n+1,m-1;t) + \gamma(n+1) P(n+1,m;t) \\
	- 
	\{ \nu + \lambda+ \mu n (1+\kappa m) +  \sigma n ( 1- \kappa m) + \gamma n \}P(n,m;t).
	\label{eq:elemaster}
\end{multline}
The joint distribution $P(n,m;t)$ that solves this equation fully characterizes the statistics of $n$ and $m$ -- proxies for the bedload flux and local bed elevation.

The average entrainment and deposition rates $E$ and $D$ over all bed elevations are $E = \lambda +\mu \langle n \rangle$ and $D=\sigma \langle n \rangle$.
The solutions of (\ref{eq:elemaster}) should be expected to adjust from whatever initial conditions toward a steady-state distribution $P_s(n,m)$ -- independent of time -- if the constant factors in the transition rates are representative of equilibrium transport conditions.
Equilibrium requires $E=D$, meaning there is no net change in elevation, and $\nu = \gamma \langle n \rangle$, meaning mass is conserved in the control volume (inflow $=$ outflow).
This Master equation describes a two-species stochastic birth-death model \citep{Cox1965} of a type well-known in population ecology \citep{Pielou1977, Swift2002} and chemical physics \citep{Gardiner1983}.
In context of this chapter, the two populations are the moving and stationary grains in the control volume.

\section{Model solutions}
\label{sec:elesolution}

Unfortunately, equation (\ref{eq:elemaster}) does not admit an analytical solution unless $\kappa=0$ (but see \citet{Swift2002} for the generating function method which fails in this case).
The difficulty originates from the product terms between $n$ and $m$ representing the bed elevation dependence of collective entrainment and deposition rates.

Because the model is Markovian, it is nevertheless possible to numerically simulate equation (\ref{eq:elemaster}) with the Gillespie algorithm \citep{Gillespie1977, Gillespie1991, Gillespie2007}. 
In conditions when the coupling constant $\kappa$ between the entrainment/deposition rates and the bed elevation is weak, it is also possible to solve the master equation approximately with mean field and Fokker-Planck approaches \citep{Haken1978,Gardiner1983}.
The Gillespie simulation algorithm is described below in \ref{sec:numerical}, while these analytical approximations are described in \ref{sec:analytical}.

@@

The Gillespie algorithm takes advantage of the defining property of a Markov process: when transition rates are independent of history, time intervals between transitions are
exponentially distributed \citep{Cox1965}.

As a result, to step the Markov process through a single transition, one can draw a first random value from the exponential distribution of transition intervals to determine the time of the next transition,
then draw a second random value to choose the type of transition that occurs using relative probabilities formed from equations (\ref{eq:rate1}-\ref{eq:rate4}). The transition is enacted by shifting $t$, $n$ and $m$ by the appropriate values to the type of transition (that is, entrainment is $m\rightarrow m-1$ and $n \rightarrow n+1$, and so on).
This procedure can be iterated to form an exact realization of the stochastic process \citep{Gillespie2007}.
The appendix chapter \ref{ch:gill} provides additional background on the stochastic simulation method.

Using this method, I simulated 4 transport conditions with 13 different values of $l$ taken across a range from $l=a$ (a single radius) to $l=10a$ (10 radii).
These values include the range exhibited by the available experimental data on bed elevation timeseries \citep{Wong2007,Singh2009,Martin2014}.

\begin{wraptable}{r}{0.6\textwidth}
	\caption{Migration, entrainment, and deposition rates at $z(m)=0$ from \citet{Ancey2008}. Units are $s^{-1}$ (probability/time). Bed elevation changes modulate these rates in accord with (\ref{eq:rate1}-\ref{eq:rate4}).}\label{tab:anceyparams}
	\begin{tabular}{cccccc} \\ 
		\toprule  
		flow & $\nu$ & $\lambda$ & $\mu$ & $\sigma$ & $\gamma$ \\
		\midrule
		(a) & 5.45  & 6.59  & 3.74 & 4.67 & 0.77 \\
		\midrule
		(g) & 7.74  & 8.42  & 4.34 & 4.95 & 0.56 \\
		\midrule
		(i) & 15.56 & 22.07 & 3.56 & 4.52 & 0.68 \\
		\midrule
		(l) & 15.52 & 14.64 & 4.32 & 4.77 & 0.48 \\
		\bottomrule
	\end{tabular}
\end{wraptable}
For the migration, entrainment, and deposition parameters representing bedload transport at each flow condition $(\nu, \lambda, \mu, \sigma, \gamma)$, I used the values measured by \citet{Ancey2008} in a series of flume experiments: these are summarized in table \ref{tab:anceyparams}.
Flow conditions are labeled (a), (g), (i), and (l), roughly in order of increasing bedload flux (see \citet{Ancey2008} for more details). 
In all simulations, I took the packing fraction $\phi = 0.6$ -- a typical value for a pile of spheres \citep[e.g.,][]{Bennett1972}, and I set $L = 22.5$cm and $a = 0.3$cm, values from the original \citet{Ancey2008} experiments.
Each simulation was run for $250$ hours of virtual time, a period selected to ensure convergence of particle activity and bed elevation statistics.

@@

\section{Discussion}
\label{sec:elediscussion}

@@

This chapter has modified the birth-death approach to describe sediment transport in a control volume. This model was applied to study the interplay between bedload transport and bed elevation fluctuations and to investigate resting time distributions of sediment undergoing burial.
This model is the first description of bedload transport and bed elevations as a coupled stochastic population model based on individual grains.
Numerical solutions and analytical approximations provided negative binomial distributions of bedload activity and normal distributions of bed elevations.
Although experiments under more natural conditions with segregation processes,  migrating bedforms, or sediment supply perturbations have shown particle activity distributions with heavier tails \citep{Dhont2018,Saletti2015} and non-Gaussian bed elevations \citep{Singh2012,Aberle2006}, our results reproduce the key features of the most controlled bed elevation \citep{Wong2007,Martin2014} and bedload transport \citep{Heyman2016,Ancey2008} experiments in the literature.

The inclusion of coupling between the bed elevation and entrainment and deposition rates revealed a novel dependence of particle activity on bed elevation changes, highlighting a new consequence of the collective entrainment process \citep{Ancey2008,Lee2018}. This coupling develops a significant variation of the particle activity moments with deviations of the bed from its mean elevation. These particle activity variations with bed elevation changes disappear in the absence of collective entrainment.

Finally, the model allowed resolution of resting times for sediment undergoing burial within the sedimentary bed. These timescales were obtained by analyzing return times from above in the bed elevation timeseries.
This analysis produces heavy-tailed power-law resting times with tail parameters sufficient to drive anomalous diffusion of bedload at long timescales.
The power-law tails were found to collapse across flow conditions using a timescale formed from the mean erosion rate and the active layer depth.
